---
title: "COVID 19 Visualization"
author: "Tanvi Hasija"
date: "2023-01-23"
output: html_document
knit: (
  function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
      encoding=encoding,
      output_file=file.path(dirname(input_file), out_dir, 'index.html')
    )
  }
 )
---

### Dataset

We will begin by using COVID-19 data from [Our World in Data](ourworldindata.org), and later, to that, we will add more data from [World Health Organization](who.int).

1.  Download [this](https://covid.ourworldindata.org/data/owid-covid-data.csv) CSV from Our World in Data.
2.  Download [this](https://covid19.who.int/WHO-COVID-19-global-data.csv) CSV from World Health Organization.

### R Packages

#### Download the required packages.

Note: Run these commands manually in the R console

    install.packages("tidyverse")
    install.packages("ggplot2")
    install.packages("ggmap")

#### Attach these packages

```{r}
library(tidyverse)
library(ggplot2)
library(ggmap)
library(scales)
```

### Phase 1: Vizualing the first dataset

#### Capturing and cleaning the data

##### Capture the World Map

```{r}
world_map <- map_data('world')
head(world_map)
```

##### Read COVID-19 CSV from OurWorldInData and understand its structure

```{r}
all_owid_cases <- read.csv('owid-covid-data.csv', header=TRUE)
names(all_owid_cases)
```

##### Group the number of cases country-wise and make sure to filter out N/A values

```{r}
country_wise_owid_cases <- all_owid_cases %>% 
  group_by(location) %>% 
  summarise(count=sum(new_cases, na.rm=TRUE)) %>%
  arrange(desc(count))
head(country_wise_owid_cases)
```

##### Filter out globally categorized data as we don't need it for our analysis

```{r}
country_wise_owid_cases <- country_wise_owid_cases[ !(country_wise_owid_cases$location %in% list("World", "High income", "Europe", "Asia", "European Union", "Upper middle income", "North America", "Lower middle income", "South America", "Low income", "Africa", "International")), ]
head(country_wise_owid_cases)
```

##### Check for the differences in country names in the COVID data and the world map data (for eg "United States" vs "USA")

```{r}
setdiff(world_map$region, country_wise_owid_cases$location)
```

##### Fix the anomallies

```{r}
country_wise_owid_cases$location[country_wise_owid_cases$location=='Congo'] = "Republic of Congo"
country_wise_owid_cases$location[country_wise_owid_cases$location=="Cote d'Ivoire"] = "Ivory Coast"
country_wise_owid_cases$location[country_wise_owid_cases$location=="Czechia"] = "Czech Republic"
country_wise_owid_cases$location[country_wise_owid_cases$location=="Democratic Republic of Congo"] = "Democratic Republic of the Congo"
country_wise_owid_cases$location[country_wise_owid_cases$location=="Eswatini"] = "Swaziland"
country_wise_owid_cases$location[country_wise_owid_cases$location=="Faeroe Islands"] = "Faroe Islands"
country_wise_owid_cases$location[country_wise_owid_cases$location=="Micronesia (country)"] = "Micronesia"
country_wise_owid_cases$location[country_wise_owid_cases$location=="Pitcairn"] = "Pitcairn Islands"
country_wise_owid_cases$location[country_wise_owid_cases$location=="Sint Maarten (Dutch part)"] = "Sint Maarten"
country_wise_owid_cases$location[country_wise_owid_cases$location=="Timor"] = "Timor-Leste"
country_wise_owid_cases$location[country_wise_owid_cases$location=="United Kingdom"] = "UK"
country_wise_owid_cases$location[country_wise_owid_cases$location=="United States"] = "USA"
country_wise_owid_cases$location[country_wise_owid_cases$location=="United States Virgin Islands"] = "Virgin Islands"

temp_ct <- country_wise_owid_cases$count[country_wise_owid_cases$location=='Antigua and Barbuda']
country_wise_owid_cases$location[country_wise_owid_cases$location=='Antigua and Barbuda'] = "Antigua"
country_wise_owid_cases[nrow(country_wise_owid_cases) + 1,] = list("Barbuda", temp_ct)

temp_ct <- country_wise_owid_cases$count[country_wise_owid_cases$location=='Bonaire Sint Eustatius and Saba']
country_wise_owid_cases$location[country_wise_owid_cases$location=='Bonaire Sint Eustatius and Saba'] = "Bonaire"
country_wise_owid_cases[nrow(country_wise_owid_cases) + 1,] = list("Sint Eustatius", temp_ct)
country_wise_owid_cases[nrow(country_wise_owid_cases) + 1,] = list("Saba", temp_ct)

temp_ct <- country_wise_owid_cases$count[country_wise_owid_cases$location=='Saint Kitts and Nevis']
country_wise_owid_cases$location[country_wise_owid_cases$location=="Saint Kitts and Nevis"] = "Saint Kitts"
country_wise_owid_cases[nrow(country_wise_owid_cases) + 1,] = list("Nevis", temp_ct)

temp_ct <- country_wise_owid_cases$count[country_wise_owid_cases$location=='Saint Vincent and the Grenadines']
country_wise_owid_cases$location[country_wise_owid_cases$location=="Saint Vincent and the Grenadines"] = "Saint Vincent"
country_wise_owid_cases[nrow(country_wise_owid_cases) + 1,] = list("Grenadines", temp_ct)

temp_ct <- country_wise_owid_cases$count[country_wise_owid_cases$location=='Trinidad and Tobago']
country_wise_owid_cases$location[country_wise_owid_cases$location=="Trinidad and Tobago"] = "Trinidad"
country_wise_owid_cases[nrow(country_wise_owid_cases) + 1,] = list("Tobago", temp_ct)
```

##### Rename the column "location" to "region" for inner join to work

```{r}
colnames(country_wise_owid_cases) <- c("region", "count")
names(country_wise_owid_cases)
```

##### Inner join the data

```{r}
data_owid <- inner_join(world_map, country_wise_owid_cases, by="region")
head(data_owid)
```

#### Plotting

##### Define a plain and aesthetics for plotting

```{r}
plain <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank(),
  panel.background = element_rect(fill = "white"),
  plot.title = element_text(hjust = 0.5)
)
aesthetics <- scale_fill_distiller(palette="Spectral", label=number_format(scale=1e-6, suffix = "M"))
```

##### Plot

```{r}
world_plot_owid <- ggplot(data = data_owid, mapping = aes(x = long, y = lat, group = group)) +
  coord_fixed(1) +
  geom_polygon(aes(fill = count)) +
  ggtitle("COVID19 Cases Worldwide") +
  plain +
  aesthetics
world_plot_owid
```

### Phase 2: Adding more data to Phase 1 dataset and vizualizing again

#### Capturing and cleaning the data

##### Read COVID-19 CSV from WorldHealthOrganization and understand its structure

```{r}
all_who_cases <- read.csv('WHO-COVID-19-global-data.csv', header=TRUE)
names(all_who_cases)
```

##### Group the number of cases country-wise and make sure to filter out N/A values

```{r}
country_wise_who_cases <- all_who_cases %>% 
  group_by(Country) %>% 
  summarise(count=sum(New_cases, na.rm=TRUE)) %>%
  arrange(desc(count))
head(country_wise_who_cases)
```

##### Check for the differences in country names in the new data and the world map data (for eg "United States" vs "USA")

```{r}
setdiff(world_map$region, country_wise_who_cases$Country)
```

##### Fix the anomallies

```{r}
country_wise_who_cases$Country[country_wise_who_cases$Country=='Bolivia (Plurinational State of)'] = "Bolivia"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Brunei Darussalam'] = "Brunei"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Cabo Verde'] = "Cape Verde"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Congo'] = "Republic of Congo"
country_wise_who_cases$Country[country_wise_who_cases$Country=="Côte d’Ivoire"] = "Ivory Coast"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Curaçao'] = "Curacao"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Czechia'] = "Czech Republic"
country_wise_who_cases$Country[country_wise_who_cases$Country=="Democratic People's Republic of Korea"] = "North Korea"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Eswatini'] = "Swaziland"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Falkland Islands (Malvinas)'] = "Falkland Islands"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Holy See'] = "Vatican"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Iran (Islamic Republic of)'] = "Iran"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Kosovo[1]'] = "Kosovo"
country_wise_who_cases$Country[country_wise_who_cases$Country=="Lao People's Democratic Republic"] = "Laos"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Micronesia (Federated States of)'] = "Micronesia"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Northern Mariana Islands (Commonwealth of the)'] = "Northern Mariana Islands"
country_wise_who_cases$Country[country_wise_who_cases$Country=='occupied Palestinian territory, including east Jerusalem'] = "Palestine"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Republic of Korea'] = "South Korea"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Republic of Moldova'] = "Moldova"
country_wise_who_cases$Country[country_wise_who_cases$Country=="Réunion"] = "Reunion"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Russian Federation'] = "Russia"
country_wise_who_cases$Country[country_wise_who_cases$Country=="Saint Barthélemy"] = "Saint Barthelemy"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Syrian Arab Republic'] = "Syria"
country_wise_who_cases$Country[country_wise_who_cases$Country=='The United Kingdom'] = "UK"
country_wise_who_cases$Country[country_wise_who_cases$Country=="Türkiye"] = "Turkey"
country_wise_who_cases$Country[country_wise_who_cases$Country=='United Republic of Tanzania'] = "Tanzania"
country_wise_who_cases$Country[country_wise_who_cases$Country=='United States of America'] = "USA"
country_wise_who_cases$Country[country_wise_who_cases$Country=='United States Virgin Islands'] = "Virgin Islands"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Venezuela (Bolivarian Republic of)'] = "Venezuela"
country_wise_who_cases$Country[country_wise_who_cases$Country=='Viet Nam'] = "Vietnam"

temp_ct <- country_wise_who_cases$count[country_wise_who_cases$Country=='Antigua and Barbuda']
country_wise_who_cases$Country[country_wise_who_cases$Country=='Antigua and Barbuda'] = "Antigua"
country_wise_who_cases[nrow(country_wise_who_cases) + 1,] = list("Barbuda", temp_ct)

temp_ct <- country_wise_who_cases$count[country_wise_who_cases$Country=='Saint Helena, Ascension and Tristan da Cunha']
country_wise_who_cases$Country[country_wise_who_cases$Country=='Saint Helena, Ascension and Tristan da Cunha'] = "Saint Helena"
country_wise_who_cases[nrow(country_wise_who_cases) + 1,] = list("Ascension Island", temp_ct)

temp_ct <- country_wise_who_cases$count[country_wise_who_cases$Country=='Saint Kitts and Nevis']
country_wise_who_cases$Country[country_wise_who_cases$Country=="Saint Kitts and Nevis"] = "Saint Kitts"
country_wise_who_cases[nrow(country_wise_who_cases) + 1,] = list("Nevis", temp_ct)

temp_ct <- country_wise_who_cases$count[country_wise_who_cases$Country=='Saint Vincent and the Grenadines']
country_wise_who_cases$Country[country_wise_who_cases$Country=="Saint Vincent and the Grenadines"] = "Saint Vincent"
country_wise_who_cases[nrow(country_wise_who_cases) + 1,] = list("Grenadines", temp_ct)

temp_ct <- country_wise_who_cases$count[country_wise_who_cases$Country=='Trinidad and Tobago']
country_wise_who_cases$Country[country_wise_who_cases$Country=="Trinidad and Tobago"] = "Trinidad"
country_wise_who_cases[nrow(country_wise_who_cases) + 1,] = list("Tobago", temp_ct)
```

##### Rename the column "Country" to "region" for inner join to work

```{r}
colnames(country_wise_who_cases) <- c("region", "count")
names(country_wise_who_cases)
```

##### Sum the metrics in the two data sets

```{r}
country_wise_owid_who_cases <- merge(country_wise_owid_cases, country_wise_who_cases, by="region")
country_wise_owid_who_cases$count <- country_wise_owid_who_cases$count.x + country_wise_owid_who_cases$count.y
head(country_wise_who_cases)
```

##### Inner join the data

```{r}
data_owid_who <- inner_join(world_map, country_wise_owid_who_cases, by="region")
head(data_owid_who)
```

#### Plotting

##### Plot

```{r}
world_plot_owid_who <- ggplot(data = data_owid_who, mapping = aes(x = long, y = lat, group = group)) +
  coord_fixed(1) +
  geom_polygon(aes(fill = count)) +
  ggtitle("COVID19 Cases Worldwide") +
  plain +
  aesthetics
world_plot_owid_who
```