---
title: "COVID 19 Visualization"
author: "Tanvi Hasija"
date: "2023-01-23"
output: html_document
knit: (
  function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
      encoding=encoding,
      output_file=file.path(dirname(input_file), out_dir, 'index.html')
    )
  }
 )
---

### Dataset

We will be using COVID-19 data from OurWorldInData.org. Download the CSV from [here](https://covid.ourworldindata.org/data/owid-covid-data.csv).

### R Packages
#### Download the required packages.
Note: Run these commands manually in the R console

```
install.packages("tidyverse")
install.packages("ggplot2")
install.packages("ggmap")
```

#### Attach these packages

```{r}
library(tidyverse)
library(ggplot2)
library(ggmap)
library(scales)
```

### Capturing and cleaning the data

#### Capture the World Map

```{r}
world_map <- map_data('world')
head(world_map)
```


#### Read COVID-19 CSV and understand its structure
```{r}
all_cases <- read.csv('owid-covid-data.csv', header=TRUE)
names(all_cases)
```

#### Group the number of cases country-wise and make sure to filter out N/A values
```{r}
country_wise_cases <- all_cases %>% 
  group_by(location) %>% 
  summarise(count=sum(new_cases, na.rm=TRUE)) %>%
  arrange(desc(count))
head(country_wise_cases)
```

#### Filter out globally categorized data as we don't need it for our analysis
```{r}
country_wise_cases <- country_wise_cases[ !(country_wise_cases$location %in% list("World", "High income", "Europe", "Asia", "European Union", "Upper middle income", "North America", "Lower middle income", "South America", "Low income", "Africa", "International")), ]
head(country_wise_cases)
```

#### Check for the differences in country names in the COVID data and the world map data (for eg "United States" vs "USA")
```{r}
setdiff(world_map$region, country_wise_cases$location)
```

#### Fix the anomallies
```{r}
country_wise_cases$location[country_wise_cases$location=='Congo'] = "Republic of Congo"
country_wise_cases$location[country_wise_cases$location=="Cote d'Ivoire"] = "Ivory Coast"
country_wise_cases$location[country_wise_cases$location=="Czechia"] = "Czech Republic"
country_wise_cases$location[country_wise_cases$location=="Democratic Republic of Congo"] = "Democratic Republic of the Congo"
country_wise_cases$location[country_wise_cases$location=="Eswatini"] = "Swaziland"
country_wise_cases$location[country_wise_cases$location=="Faeroe Islands"] = "Faroe Islands"
country_wise_cases$location[country_wise_cases$location=="Micronesia (country)"] = "Micronesia"
country_wise_cases$location[country_wise_cases$location=="Pitcairn"] = "Pitcairn Islands"
country_wise_cases$location[country_wise_cases$location=="Sint Maarten (Dutch part)"] = "Sint Maarten"
country_wise_cases$location[country_wise_cases$location=="Timor"] = "Timor-Leste"
country_wise_cases$location[country_wise_cases$location=="United Kingdom"] = "UK"
country_wise_cases$location[country_wise_cases$location=="United States"] = "USA"
country_wise_cases$location[country_wise_cases$location=="United States Virgin Islands"] = "Virgin Islands"

temp_ct <- country_wise_cases$count[country_wise_cases$location=='Antigua and Barbuda']
country_wise_cases$location[country_wise_cases$location=='Antigua and Barbuda'] = "Antigua"
country_wise_cases[nrow(country_wise_cases) + 1,] = list("Barbuda", temp_ct)

temp_ct <- country_wise_cases$count[country_wise_cases$location=='Bonaire Sint Eustatius and Saba']
country_wise_cases$location[country_wise_cases$location=='Bonaire Sint Eustatius and Saba'] = "Bonaire"
country_wise_cases[nrow(country_wise_cases) + 1,] = list("Sint Eustatius", temp_ct)
country_wise_cases[nrow(country_wise_cases) + 1,] = list("Saba", temp_ct)

temp_ct <- country_wise_cases$count[country_wise_cases$location=='Saint Kitts and Nevis']
country_wise_cases$location[country_wise_cases$location=="Saint Kitts and Nevis"] = "Saint Kitts"
country_wise_cases[nrow(country_wise_cases) + 1,] = list("Nevis", temp_ct)

temp_ct <- country_wise_cases$count[country_wise_cases$location=='Saint Vincent and the Grenadines']
country_wise_cases$location[country_wise_cases$location=="Saint Vincent and the Grenadines"] = "Saint Vincent"
country_wise_cases[nrow(country_wise_cases) + 1,] = list("Grenadines", temp_ct)

temp_ct <- country_wise_cases$count[country_wise_cases$location=='Trinidad and Tobago']
country_wise_cases$location[country_wise_cases$location=="Trinidad and Tobago"] = "Trinidad"
country_wise_cases[nrow(country_wise_cases) + 1,] = list("Tobago", temp_ct)
```

#### Rename the column "location" to "region" for inner join to work
```{r}
colnames(country_wise_cases) <- c("region", "count")
names(country_wise_cases)
```

#### Inner join the data
```{r}
data <- inner_join(world_map, country_wise_cases, by="region")
head(data)
```

#### Define a plain and aesthetics for plotting
```{r}
plain <- theme(
  axis.text = element_blank(),
  axis.line = element_blank(),
  axis.ticks = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  axis.title = element_blank(),
  panel.background = element_rect(fill = "white"),
  plot.title = element_text(hjust = 0.5)
)
aesthetics <- scale_fill_distiller(palette="Spectral", label=number_format(scale=1e-6, suffix = "M"))
```

#### Plot
```{r}
world_plot <- ggplot(data = data, mapping = aes(x = long, y = lat, group = group)) +
  coord_fixed(1) +
  geom_polygon(aes(fill = count)) +
  ggtitle("COVID19 Cases Worldwide") +
  plain +
  aesthetics
world_plot
```
